% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage[margin=0.9in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother





\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Executive Summary: Cell-Type-Specific Biomarkers in Human Liver Fibrosis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Executive Summary: Cell-Type-Specific Biomarkers in Human Liver
Fibrosis}
\author{}
\date{2026-06-07}
\begin{document}
\maketitle


\subsection{Objective}\label{objective}

Discover and prioritize cell-type-specific biomarkers and therapeutic
targets in human liver fibrosis from single-cell transcriptomics
(primary: GSE136103, cirrhotic vs healthy; validation: GSE244832,
MASLD/MASH F0 to F4), focusing on the hepatic stellate/myofibroblast,
macrophage/monocyte, and endothelial compartments.

\subsection{Approach}\label{approach}

A reproducible Snakemake + Scanpy pipeline ingests and QCs the data,
integrates over donor (not disease) and verifies the fibrosis signal is
preserved, annotates and validates the three disease compartments, runs
differential expression, characterizes mechanism (pathway/TF activity
and ligand-receptor analysis with downstream-target corroboration), and
prioritizes candidates with a transparent composite score (DE effect,
cell-type specificity, cross-dataset reproducibility, druggability,
biomarker accessibility, and an explainable XGBoost/SHAP signal). Two
choices keep the ranked list trustworthy rather than a marker
re-discovery: differential expression is donor-aware pseudobulk (the
replication unit is the donor, not the cell, which avoids the
pseudoreplication that inflates cell-level tests), and the composite is
gated so a gene must be statistically DE-significant (at the compartment
or disease-subpopulation level) and cell-type-specific before
specificity, druggability, or accessibility can lift it; abundant
lineage markers and ambient cross-compartment leakage are filtered out.

\subsection{Top candidates}\label{top-candidates}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lllllll@{}}
\toprule\noalign{}
& rank & gene & compartment & composite & accessibility\_class &
druggability \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
0 & 1.0 & PDGFRA & stellate\_myofibroblast & 0.671 & surface & 0.90 \\
1 & 2.0 & ACKR1 & endothelial & 0.635 & surface & 0.30 \\
2 & 3.0 & LUM & stellate\_myofibroblast & 0.624 & secreted & 0.10 \\
3 & 4.0 & DCN & stellate\_myofibroblast & 0.612 & secreted & 0.10 \\
4 & 5.0 & TREM2 & macrophage\_monocyte & 0.606 & surface & 0.60 \\
5 & 6.0 & PLVAP & endothelial & 0.581 & surface & 0.40 \\
7 & 7.0 & COL3A1 & stellate\_myofibroblast & 0.553 & secreted & 0.20 \\
9 & 8.0 & MMP7 & cholangiocyte & 0.545 & intracellular & 0.10 \\
10 & 9.0 & SPP2 & cholangiocyte & 0.544 & intracellular & 0.10 \\
11 & 10.0 & MMP2 & stellate\_myofibroblast & 0.542 & secreted & 0.55 \\
\end{longtable}
}

Top 10 shown. The full ranked 20, with per-candidate rationale,
component scores, and significance/specificity flags, is in
\texttt{results/08\_score/candidate\_scores.tsv} and the full report.

\subsection{Interpretation}\label{interpretation}

\textbf{Hepatic stellate / myofibroblast} (PDGFRA, LUM, DCN, COL3A1,
MMP2) are ECM and PDGF-axis genes consistent with activated-HSC /
myofibroblast biology. \textbf{PDGFRA} is the stand-out
\textbf{therapeutic} target: a cell-surface receptor tyrosine kinase
already druggable with approved TKIs (imatinib / nilotinib class). The
secreted matrix proteins (LUM, DCN, COL3A1, MMP2) are
\textbf{diagnostic}-accessible: type III collagen turnover (COL3A1)
underlies the clinically validated circulating PRO-C3 fibrosis marker,
connecting this arm directly to a non-invasive readout.

\textbf{Scar-associated macrophages.} \textbf{TREM2} marks the CD9 /
SPP1 / TREM2 pro-fibrotic macrophage program defined in this primary
cohort (Ramachandran et al.~2019, \emph{Nature}). It is dual-use: a
\textbf{therapeutic} target (surface receptor, agonist antibodies in
clinical development) and a \textbf{diagnostic} one (soluble TREM2 is
measurable in blood), the strongest single pick for tracking the
fibrotic macrophage niche.

\textbf{Scar endothelium.} \textbf{PLVAP} and \textbf{ACKR1} flag the
disease-associated (capillarized) endothelial population, also from
Ramachandran 2019; these read as mechanistic / imaging markers rather
than direct drug targets.

\textbf{Ductular reaction.} \textbf{MMP7} and \textbf{SPP2} mark the
cholangiocyte / biliary response; circulating MMP-7 is an established
fibrosis / cholestasis biomarker, supporting MMP7 as a
\textbf{diagnostic} candidate.

The most actionable picks are therefore \textbf{PDGFRA} (druggable
target), \textbf{TREM2} (dual diagnostic / target for the macrophage
niche), and the \textbf{secreted collagen / proteoglycan panel} (COL3A1,
LUM, DCN) as circulating-biomarker candidates.

\textbf{Calibration.} Against a 37-gene panel of established
human-liver-fibrosis markers, the pipeline recovers 10 as headline
candidates and surfaces 25 as scored candidates, so the list's
sensitivity is moderate and the gaps are cohort-driven (small-n dilution
of subset markers) rather than scoring artifacts.

\textbf{Limitations.} The primary cohort is small (5 cirrhotic vs 5
healthy donors; 3 vs 3 for the stellate and cholangiocyte compartments),
and cross-dataset reproducibility against the MASLD/MASH validation set
is weak (rank correlation near zero), driven by the scRNA vs snRNA
modality and disease-context mismatch rather than by the primary signal;
confidence is therefore strongest for the well-powered,
internally-significant stellate and macrophage hits.

\textbf{Full report and source code:}
\url{https://github.com/ericmalekos/sc-liver}




\end{document}
